---
title: "Harvardx Data Science Choose-Your-Own-Project"
author: "Da Xuan"
date: "6/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction of dataset and purpose of current project

This data is downloaded from Kaggle (https://www.kaggle.com/abcsds/pokemon).
The data set includes 800 observations with data including 1) Pokemon name, 2) Type 1 Element, 3) Type 2 Element, 4) Total Stats, 5) Health Point, 6) Attack point, 7) Defense point, 8) Special Attack point, 9) Special Defense point, 10) Speed point, 11) Generation, and 12) Legendary class. 


The data consists of ALL the pokemons and their stats. That is, the data is not a 'sample' of a population; the data includes all observation in the population.

The task is to correctly classify the types of pokemon using Machine Learning models based on the variables in the dataset. 


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
