---
title: "Harvardx Data Science Choose-Your-Own-Project"
author: "Da Xuan"
date: "6/8/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

This data is downloaded from Kaggle (https://www.kaggle.com/abcsds/pokemon).
The data set includes 800 observations with data including 1) Pokemon name (Name), 2) Type 1 Element (`Type 1`), 3) Type 2 Element (`Type 2`), 4) Total Stats (Total), 5) Health Point (HP), 6) Attack point (Attack), 7) Defense point (Defense), 8) Special Attack point (Sp. Atk), 9) Special Defense point (Sp. Def), 10) Speed point (Speed), 11) Generation (Generation), and 12) Legendary class (Legendary). 


It is to note that the data consists of ALL the pokemons and their stats. That is, the data is not a 'sample' of a population; the data includes all observation in the population. The task is 1) to create a linear combination of variables to predict Pokemon Types, and 2) determine the structure and latent variables of the dataset by creating linear combinations of variables (ie., reduce the dimensions of the dataset). 

The key steps taken to complete the tasks were: 1) exploring the dataset and understanding the data structure, 2) exploring the relationship between the variables, 3) selecting the best model for predicting Pokemon's types (ie., 'Type 1`), and 4) obtaining a principal components analysis of the data.

The following are the libraries used in exploring the data. 


```{r libraries}


```

## DATA PREPARATION

* Download data
```{r data source}
url <- "https://raw.githubusercontent.com/dxng-sg/HarvardxPokemon/master/datasets_121_280_Pokemon.csv"
dl <- tempfile()
download.file(url, dl)
dat <- read_csv(dl)
file.remove(dl)

```


## DATA DESCRIPTION

* Exploring the dataset

```{r structure, tidy=TRUE, message=FALSE, warning = FALSE}
## To identify the data type of each variables (ie., factor, numeric, or character)
head(dat)
```


